\propertitle

### 2019-11-15 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-14T22:48:47+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

We want to stress the usefulness of making a quantified guess about the
activity of a larger brain area. Such a guess seems indeed to be the
primary idea behind recording a sample from that area. %  Second, it allows us
% to make comparisons across experimental sessions; such comparisons would me
% difficult or meaningless if made with the recorded samples, which generally
% comprise non-overlapping sets of neurons and differ in size.
There's also an important methodological reason. Maximum-entropy methods
have been used to assess the informational sufficiency of pairwise
correlations versus correlations of higher order \citep[see for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
But when the difference in size between sample and full network is too
large, correlation sufficiency for the sample implies the \emph{lack} of
correlation sufficiency for the larger network, and vice versa, as will be
explained in detail in \sect~\ref{sec:marginalization}. Therefore,
maximum-entropy applications at the sample level can be deceptive for
questions of statistical sufficiency, and a full-network application is
more reliable. We show this point in \sect~\ref{sec:application} for a real
data set.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb
To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the full-network
distributions, for $N=10\,000$, obtained constraining the first two moments
(equivalent to constraining means and correlations) and the first four
moments. The frequency distribution of the sample is also shown for
comparison. In both data sets the two-moment constraint leads to a
distribution quite different from the four-moment constraint. For the first
data set in particular, one distribution is bimodal, the other unimodal:
this difference shows that the first two moments alone are \emph{not}
informationally sufficient for this specific data set.

It is instructive to contrast the distributions for the full network with
those obtained applying the maximum-entropy method \emph{at the sample
  level} with the same constraints, as traditionally done. The sample-level
distributions are shown as an inset in the plot for the first data set. We
see that constraining four moments doesn't lead to a distribution much
different from constraining only two. This would suggest that two moments
are informationally almost sufficient. This conclusion would be deceptive:
it obviously cannot be extended to the full network. We discuss this point
more in detail in the next section.

\section{Example application for sufficiency of partial statistics:\\the
  issue with methods at the sample level}
\label{sec:marginalization}

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been applied to questions
of sufficient statistics. For example: do pairwise correlations in the
neuronal activity provide the same information as the full frequency
distribution of activities? In this section we discuss how our proposed
application bears on this kind of questions. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our discussion
% holds for any sets of statistics.

The question must be made more precise first. We want to assess our degree
of belief about the frequencies of the activities of the sample, or about
the activity of the sample in a new time bin. For this assessment we should
use all measured data we have. But it sometimes happens that dropping part
of the data -- for example, the measured third- and higher-order moments --
leaves our assessment almost unchanged: the dropped data are
\emph{informationally irrelevant}, or almost so. The remaining data -- for
example, first and second moments -- are then \emph{informationally
  sufficient}. % This informational quasi-sufficiency is interesting because
% it may give clues about network organization and dynamics.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/trunct_comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
% comparison_ME_sample_full_levels.nb
%
% relentropies of sample2mom, marg2mom, sample4mom, marg4mom:
% 322.72, -85.7627, -42.8005, -4.23746
% correspond to probability ratios of
% 1., 8.11718\times 10^{102}, 3.69524\times 10^{121}, 2.06704\times 10^{138}
% comparison_ME_sample_full_levels.nb
%
But does the application at the full-network level lead to appreciably
different results from that at the sample level? after all we're interested
in an approximate informational sufficiency, not in an analytically exact
one. A correct answer can only be given case by case.
Figure~\ref{fig:comparisons_marginals} gives a graphical answer to this
question in the case of our first data set. The upper plot shows the
maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{\textbf{--~--~--}}) and
from four moments (\textcolor{mypurpleblue}{\textbf{------}}). The lower
plot shows the maximum-entropy distributions for the sample, from two
moments (\textcolor{mygreen}{$\bm{\triangledown}$}) and from four moments
(\textcolor{mypurpleblue}{$\bm{\square}$}), and the distributions obtained
by marginalizing the full-network distribution to the sample size, from two
moments (\textcolor{mygreen}{$\bm{\blacktriangledown}$}) and from four
moments (\textcolor{mypurpleblue}{$\bm{\blacksquare}$}). The measured
frequency distribution (\textcolor{myred}{\LARGE$\bm{\bullet}$}) is also
shown. We note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two sets of statistics are not even approximately equivalent
  for inferential purposes.
\item The application at the sample level leads to deceivingly similar
  distributions (\textcolor{mygreen}{$\bm{\triangledown}$} and
  \textcolor{mypurpleblue}{$\bm{\square}$}), which can only be distinguished on a
  logarithmic scale. This could lead to the erroneous conclusion that the
  two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the maximum-entropy application at the sample level and at the
full-network level lead to different results; the latter is not only more
meaningful but also superior because it shows clearly the different
informational sufficiency of the two sets of statistics.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-15T15:15:50+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

We want to stress the usefulness of making a quantified guess about the
activity of a larger brain area. Such a guess seems indeed to be the
primary idea behind recording a sample from that area. %  Second, it allows us
% to make comparisons across experimental sessions; such comparisons would me
% difficult or meaningless if made with the recorded samples, which generally
% comprise non-overlapping sets of neurons and differ in size.
There's also an important methodological reason. Maximum-entropy methods
have been used to assess the informational sufficiency of pairwise
correlations and correlations of higher order \citep[see for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
But when the difference in size between sample and full network is too
large, correlation sufficiency for the sample implies the \emph{lack} of
correlation sufficiency for the larger network, and vice versa. Therefore,
maximum-entropy applications at the sample level can be deceptive for
questions of statistical sufficiency, and a full-network application is
more reliable. We discuss this point in detail in \sect~\ref{sec:marginalization}.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb
To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the full-network
distributions, for $N=10\,000$, obtained constraining the first two moments
(equivalent to constraining means and correlations) and the first four
moments. The frequency distribution of the sample is also shown for
comparison. In both data sets the two-moment constraint leads to a
distribution quite different from the four-moment constraint. For the first
data set in particular, one distribution is bimodal, the other unimodal.
This visual difference shows directly -- without the need to calculate
relative entropies, for example -- that the first two moments alone are
\emph{not} informationally sufficient for this specific data set.

It is instructive to contrast the conclusions about correlation sufficiency
reached through this full-network analysis, with those we would have
reached by applying the maximum-entropy method \emph{at the sample
  level}, as traditionally done. Such a comparison involves an important
methodological caveat, so we discuss it in full in the next section.

\section{Quantifying the importance of higher-order correlations:\\
  the limitations of methods at the sample level}
\label{sec:marginalization}

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep{gersteinetal1985} or
\enquote{interaction}
\citep{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep{bohteetal2000,amarietal2003} of neuronal
activity. In this section we discuss how our proposed
application bears on this kind of quantification. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

Let us translate this question into more precise terms first, because
\enquote{cooperativity}, \enquote{interaction}, and similar notions are
vague. A translation in terms of \emph{informational sufficiency} seems to
suit all these notions \parentext{for an intuitive understanding of
  sufficiency, which goes back to \cite{fisher1922}, see \eg\
  \cite[\sect~4.5]{bernardoetal1994}; \cite[\chap~8,
  \sect~14.2]{jaynes1994_r2003}; \cite{cifarellietal1982};
  \cite{kullbacketal1951}. For technical results and connections with the
  notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,cifarellietal1980,cifarellietal1981,diaconisetal1981,lauritzen1982_r1988,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,lauritzen2007,ayetal2015}}.
The idea is as follows. We want to assess our degree of belief about the
frequencies of the activities of the sample, or about the activity of the
sample in a new time bin. For this assessment we should use all measured
data we have. But it sometimes happens that dropping part of the data --
for example, the measured third- and higher-order moments -- leaves our
assessment almost unchanged: the dropped data are \emph{informationally
  irrelevant}, or almost so. The remaining data -- for example, first and
second moments -- are then \emph{informationally
  sufficient}. % This informational quasi-sufficiency is interesting because
% it may give clues about network organization and dynamics.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/trunct_comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
% comparison_ME_sample_full_levels.nb
%
% relentropies of sample2mom, marg2mom, sample4mom, marg4mom:
% -322.72, -85.7627, -42.8005, -4.23746 nats =
% -1401.56, -372.462, -185.88, -18.403 db
% correspond to probability ratios of
% 1., 8.11718\times 10^{102}, 3.69524\times 10^{121}, 2.06704\times 10^{138}
%
% difference in relentropies between 4 mom and 2 mom:
% sample level: 279.92 nats = 1215.68 db
% marg level: 81.5252 nats =  354.059 db
% correspond to prob ratios of
% 3.69524\times 10^{121},2.5465\times 10^{35}
%
% comparison_ME_sample_full_levels.nb
%
But does the application at the full-network level lead to appreciably
different results from that at the sample level? after all we're interested
in an approximate informational sufficiency, not in an analytically exact
one. A correct answer can only be given case by case.
Figure~\ref{fig:comparisons_marginals} gives a graphical answer to this
question in the case of our first data set. The upper plot shows the
maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{\textbf{--~--~--}}) and
from four moments (\textcolor{mypurpleblue}{\textbf{------}}). The lower
plot shows the maximum-entropy distributions for the sample, from two
moments (\textcolor{mygreen}{$\bm{\triangledown}$}) and from four moments
(\textcolor{mypurpleblue}{$\bm{\square}$}), and the distributions obtained
by marginalizing the full-network distribution to the sample size, from two
moments (\textcolor{mygreen}{$\bm{\blacktriangledown}$}) and from four
moments (\textcolor{mypurpleblue}{$\bm{\blacksquare}$}). The measured
frequency distribution (\textcolor{myred}{\LARGE$\bm{\bullet}$}) is also
shown. We note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two sets of statistics are not even approximately equivalent
  for inferential purposes.
\item The application at the sample level leads to deceivingly similar
  distributions (\textcolor{mygreen}{$\bm{\triangledown}$} and
  \textcolor{mypurpleblue}{$\bm{\square}$}), which can only be distinguished on a
  logarithmic scale. This could lead to the erroneous conclusion that the
  two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the maximum-entropy application at the sample level and at the
full-network level lead to different results; the latter is not only more
meaningful but also superior because it shows clearly the different
informational sufficiency of the two sets of statistics.
